IPv6 over Low-Power Wireless Personal Area Networks\\ MSc. Thesis Proposal

### 2010-01-12 doc/msc-proposal/msc-proposal.tex
notepad

\large Supervisor: J\"urgen Shoenw\"alder\\[0.5cm]
{\large \today}

@@

\newpage
\tableofcontents
\listoffigures 
\newpage

@@

\chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks. The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes). Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.

@@

An 802.15.4 network may operate in either the star or the peer-to-peer topology. In the star topology devices communicate with a single central                           a personal area network (PAN) coordinator. In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another. Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology. However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications. All devices operating on a network of either topology have unique 64 bit extended addresses. This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.

@@

%%
%	Transmission of IPv6 Packets
%%
\chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard. The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other. On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.

@@

6lowpan (Internet Protocol Version 6 (IPv6) over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks. The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4. The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.

@@

6LoWPAN supports stateless address autoconfiguration, which allows to obtain the Interface Identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE802.15.4 device. Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.  In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address and the interface identifier is formed from this 48-bit address:
\begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}

The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the Interface Identifier to the prefix \texttt{FE80::/64}.

@@

\section{Adaptation Layer}\label{sec:adapt.layer}
The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets. As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.  Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer. Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size. Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.

Even though 802.15.4 networks are expected to commonly use mesh routing, the IEEE 802.15.4-2003 \cite{ieee802.15.4} specification does not define such capability. Mesh routing allows two devices  without direct reachability to communicate with each other. In such mesh scenarios, intermediate devices are used as forwarders towards the final destination. In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses the originator and final destination addresses need to be known. For this purpose a Mesh Addressing header is introduced.

IPv6 datagrams transported over IEEE 802.15.4 are prefixed by an encapsulation header stack. Each header in the header stack contains a header type followed by zero or more header fields.  There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header, Fragmentation Header and Dispatch Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively). These headers must appear in the specified order. First three of them are optional, whereas the last one is required.

@@

The first and subsequent fragment header types are specified by the binary sequences \texttt{11000} and \texttt{11100} respectively.  The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.  The \texttt{datagram\_tag} field is the same for all link fragments of a datagram and incremented by the sender for successive fragmented datagrams. The \texttt{datagram\_offset} field is present only in the second and subsequent link fragments and specifies the offset of the fragment from the beginning of the payload datagram.

\subsection{Broadcast Header}\label{subsec:broad.header}
Broadcast is an additional mesh routing functionality. For that a broadcast header is defined the format of which is shown in Figure \ref{fig:broad.header}.
\begin{figure}[htp]
\begin{mylisting}
\begin{verbatim}
                                       1
                   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
                  |0|1|0|1|0|0|0|0|Sequence Number|
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Broadcast Header
}\label{fig:broad.header}
\end{figure}

@@

\section{Header Compression}\label{sec:hc}
As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6. The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.  The latter uses 8 octets in the header, which leaves 33 octets for application data. 
The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.

RFC 4944 \cite{rfc4944} defines stateless header compression mechanisms for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers. However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses derived directly from IEEE 802.15.4 addresses. When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line. Internet Draft document \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts. In addition, it also defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.

\subsection{HC1 Header Encoding}
The HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link. The following IPv6 header values are expected to be common on 6LoWPAN network: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP. The Hop Limit (8 bits) field is the only one that always needs to be carried. Using the HC1 encoding allows to compress a 40 byte IPv6 header to 2 bytes.

@@

\subsection{IPHC Header Encoding}
The IPHC encoding format enables effective compression of the IPv6 header relying on information pertaining to the entire 6LoWPAN network. The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an IID derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses. The compression mechanism is adapted for these values of the header fields and in such scenario the LOWPAN IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC encoding) with link-local communication.  When routing over multiple IP hops, IPHC can compress the IPv6 header down to 7 octets (1-octet dispatch, 1-octet IPHC encoding, 1-octet Hop Limit, 2-octet Source Address, and 2-octet Destination Address).  In this case, the Hop Limit may not be compressed because it needs to be decremented at each hop and stateful address compression is applied to the source and destination IPv6 addresses because they do not statelessly match the source and destination link layer addresses on intermediate hops.

The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$ and a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields. The IPHC encoding may be extended by another octet to support additional contexts.

@@

Stateful address compression expects that a conceptual context is shared between the node that compresses a packet and the node(s) that need to expand it.  However, the Internet Draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context. If the Context Identifier (CID) field (see Figure \ref{fig:iphc.format}) is set to 1 then an additional 8-bit Context Identifier Extension field is used and immediately follows the IPHC Base Encoding. This field identifies the pair of contexts used for the IPv6 source and destination address compression. The context identifiers are 4 bits long.

@@

\subsubsection{IPv6 Extension Header Compression}
The NHC encoding for IPv6 Extension Headers (shown in Figure \ref{fig:ipv6.ext.header}) is composed of a single octet followed by the IPv6 Extension Header.  IPv6 Extension Headers use  $1110$ bit pattern as bits $0$ through $3$ in the NHC encoding. The latter 3 bits represent an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet. Currently, the following values of the EID field have been specified:  $0$ -- IPv6 Hop-by-Hop Options Header \cite{rfc2460}; $1$ -- IPv6 Routing Header \cite{rfc2460};  $2$ -- IPv6 Fragment Header \cite{rfc2460}; $3$ -- IPv6 Destination Options Header \cite{rfc2460}; $4$ -- IPv6 Mobility Header \cite{rfc3775}; $5$ -- Reserved; $6$ -- Reserved; $7$ -- IPv6 Header. The last bit of the NHC encoding indicates whether the NHC compression is used for the following header.

\begin{figure}[htp]
\begin{mylisting}
\begin{verbatim}
                0   1   2   3   4   5   6   7
              +---+---+---+---+---+---+---+---+
              | 1 | 1 | 1 | 0 |    EID    |NH |
              +---+---+---+---+---+---+---+---+
\end{verbatim}
\end{mylisting}
\caption{IPv6 Extension Header Encoding}\label{fig:ipv6.ext.header}
\end{figure}

@@

\nocite{ieee802.15.4} 
\nocite{eui64} 
\nocite{rfc780}
\nocite{rfc2460}
\nocite{rfc2464}
\nocite{rfc2474}
\nocite{rfc3168}
\nocite{rfc3775}
\nocite{rfc4291} 
\nocite{rfc4443}
\nocite{rfc4861}
\nocite{rfc4862}
\nocite{rfc4919}
\nocite{rfc4944} 
\nocite{draft-usecases-05} 
\nocite{draft-hc-06} 
\nocite{draft-nd-07}

>>>>

\large Supervisor: J\"urgen Sh\"onw\"alder\\[0.5cm]
{\large \today}

@@

\newpage
\tableofcontents
%\listoffigures
\newpage

@@

\chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks providing nodo-to-node frame dilivery between devices within reachable distance from each other. Such networks are typically limited to a personal operating space of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.

@@

An 802.15.4 network may operate in either the star or the peer-to-peer topology. In the star topology devices communicate with a single central                            personal area network (PAN) coordinator. In the peer-to-peer topology a device can communicate with any other device as long as they are in range of one another. Based on the peer-to-peer topology more complex network formations may be constructed, such as mesh networking topology. However, the standard does not define a network layer, therefor routing is not directly supported, but such an additional layer can add support for multihop communications. All devices operating on a network of either topology have unique 64 bit extended addresses. This address can be used for direct communication within the PAN, or it can be exchanged for a short 16 bit address allocated by the PAN coordinator.

@@

%%
%   Transmission of IPv6 Packets
%%
\chapter{Transmission of IPv6 Packets over 802.15.4}\label{ch:6lowpan}
As described in Chapter \ref{ch:ieee802.15.4}, the IEEE 802.15.4-2003 standard specifies the physical layer and media access control for low-rate wireless personal area network providing node-to-node frame delivery between the devices within reachable distance from each other. On top of IEEE 802.15.4, a network layer (layer 3 of the seven-layer OSI model) has to be defined that provides end-to-end packet delivery including routing through intermediate nodes.

@@

The 6lowpan (Internet Protocol Version 6 (IPv6) over Low power Wireless Personal Area Networks) working group of the IETF is concerned with the specification of mechanisms to allow IPv6 packet transmission over IEEE 802.15.4 based networks. The working group has already completed two documents: "IPv6 over Low-Power Wireless Personal Area Networks (6LoWPANs): Overview, Assumptions, Problem Statement, and Goals" \cite{rfc4919} that documents and discusses the problem statement and "Transmission of IPv6 Packets over IEEE 802.15.4 Networks" \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4. The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.

@@

6LoWPAN supports stateless address autoconfiguration, which allows to obtain the interface identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE 802.15.4 device. Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.  In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address. The interface identifier is formed from this 48-bit address as defined in the "IPv6 over Ethernet" specification \cite{rfc2464}:
\begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}

The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the interface identifier to the prefix \texttt{FE80::/64}.

@@

\section{Adaptation Layer}\label{sec:adapt.layer}
The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets. As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.  Taking into account the maximum frame overhead of 25 octets (disabled security), 102 octets are left at the media access control layer. Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size. Therefore, a fragmentation and reassembly adaptation layer is provided at the layer below IP.

Even though 802.15.4 networks are expected to commonly use mesh routing, the IEEE 802.15.4-2003 \cite{ieee802.15.4} specification does not define such capability. Mesh routing allows two devices  without direct reachability to communicate with each other. In such mesh scenarios, intermediate devices are used as forwarders towards the final destination. In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses, the originator and final destination addresses need to be known. For this purpose a Mesh Addressing header is introduced.

IPv6 datagrams transported over IEEE 802.15.4 are prefixed by an encapsulation header stack. Each header in the header stack contains a header type followed by zero or more header fields.  There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header, Fragmentation Header and Dispatch Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively). These headers must appear in the specified order. The first three of them are optional, whereas the last one is required.

@@

The first and subsequent fragment header types are identified by the binary sequences \texttt{11000} and \texttt{11100} respectively.  The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.  The \texttt{datagram\_tag} field is the same for all link fragments of a datagram and incremented by the sender for successive fragmented datagrams. The \texttt{datagram\_offset} field is present only in the second and subsequent link fragments and specifies the offset of the fragment from the beginning of the payload datagram.

\subsection{Broadcast Header}\label{subsec:broad.header}
Broadcast is an additional mesh routing functionality. For that a broadcast header is defined, the format of which is shown in Figure \ref{fig:broad.header}.
\begin{figure}[htp]
\begin{mylisting}
\begin{verbatim}
                                       1
                   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
                  |0|1|0|1|0|0|0|0|Sequence Number|
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Broadcast Header
}\label{fig:broad.header}
\end{figure}

@@

\section{Header Compression}\label{sec:hc}
As described in Section \ref{sec:adapt.layer}, 81 octets are left in an IEEE 802.15.4 frame for IPv6. The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.  The latter uses 8 octets in the header, which leaves 33 octets for application data.
The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet. This makes header compression almost unavoidable.

RFC 4944 \cite{rfc4944} defines a stateless header compression mechanism for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers. However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses can be derived directly from IEEE 802.15.4 addresses. When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line. The internet draft \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts. In addition, it defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.

\subsection{HC1 Header Encoding}

Devices in the same 6LoWPAN network share some state.  This makes it possible to compress headers without explicitly building any compression context state.  HC1 header compression makes use of this. The following IPv6 header values are expected to be common on 6LoWPAN networks: Version is IPv6; both IPv6 source and destination addresses are link local;  the IPv6 interface identifiers for the source or destination addresses can be inferred from the layer two source and destination addresses; the packet length can be inferred either from layer two or from the "datagram\_size" field in the fragment header; both the Traffic Class and the Flow Label are zero; and the Next Header is UDP, ICMP or TCP. The Hop Limit (8 bits) field is the only one that always needs to be carried. Using the HC1 encoding allows to compress a 40 byte IPv6 header to 2 bytes.

@@

\subsection{IPHC Header Encoding}
The IPHC encoding format enables effective compression of the IPv6 header relying on information pertaining to the entire 6LoWPAN network. The encoding assumes the following common case parameters for 6LoWPAN communication: Version is $6$; Traffic Class and Flow Label are both zero; Payload Length can be inferred from lower layers from either the 6LoWPAN Fragmentation header or the IEEE 802.15.4 header; Hop Limit will be set to a well-known value by the source; addresses assigned to 6LoWPAN interfaces will be formed using the link-local prefix or a single routable prefix assigned to the entire 6LoWPAN network; addresses assigned to 6LoWPAN interfaces are formed with an interface identifier derived directly from either the 64-bit extended or 16-bit short IEEE 802.15.4 addresses. The compression mechanism is adapted for these values of the header fields and in such scenario the LOWPAN IPHC can compress the IPv6 header down to two octets (1-octet dispatch, 1-octet LOWPAN\_IPHC encoding) with link-local communication.  When routing over multiple IP hops, IPHC can compress the IPv6 header down to 7 octets (1-octet dispatch, 1-octet IPHC encoding, 1-octet Hop Limit, 2-octet Source Address, and 2-octet Destination Address).  In this case, the Hop Limit may not be compressed because it needs to be decremented at each hop and stateful address compression is applied to the source and destination IPv6 addresses because they do not statelessly match the source and destination link layer addresses on intermediate hops.

The IPv6 header encrypted with the IPHC encoding format, as shown in Figure \ref{fig:iphc.header}, consists of a 3 bit dispatch value, $110$ and a 13 bit IPHC base encoding (shown in Figure \ref{fig:iphc.format}), describing how the IPv6 header is compressed, followed by compressed IPv6 header fields. The IPHC encoding may be extended by another octet to support additional contexts for stateful address compression described later in this Section.

@@

Stateful address compression expects that a conceptual context is shared between the node that compresses a packet and the node(s) that need to expand it.  However, the internet draft \cite{draft-hc-06} does not specify how the contexts are shared, maintained and what information is contained within a context. If the Context Identifier (CID) field (see Figure \ref{fig:iphc.format}) is set to 1 then an additional 8-bit Context Identifier Extension field is used and immediately follows the IPHC Base Encoding. This field identifies the pair of contexts used for the IPv6 source and destination address compression. The context identifiers are 4 bits long.

@@

\subsubsection{IPv6 Extension Header Compression}
The NHC encoding for IPv6 Extension Headers (shown in Figure \ref{fig:ipv6.ext.header}) is composed of a single octet followed by the IPv6 Extension Header.  IPv6 Extension Headers use  $1110$ bit pattern as bits $0$ through $3$ in the NHC encoding. The latter 3 bits represent an IPv6 Extension Header ID (EID) which identifies the IPv6 Extension Header immediately following the NHC octet. The defined values of the EID field are shown in Table \ref{table:ipv6.ext.header}. 
\begin{table}[htp]
\begin{center}
        \begin{tabular}{|c|l|}
          \hline
          EID & Description\\
          \hline
          \hline          
           0 & IPv6 Hop-by-Hop Options Header \cite{rfc2460}\\
           1 & IPv6 Routing Header \cite{rfc2460}\\
           2 & IPv6 Fragment Header \cite{rfc2460}\\
           3 & IPv6 Destination Options Header \cite{rfc2460}\\
           4 & IPv6 Mobility Header \cite{rfc3775}\\
           5 & Reserved\\
           6 & Reserved\\
           7 & IPv6 Header\\
          \hline
        \end{tabular}
\end{center}
\caption{IPv6 Extension Header ID values}\label{table:ipv6.ext.header}
\end{table}

The last bit of the NHC encoding indicates whether the NHC compression is used for the following header.
\begin{figure}[htp]
\begin{mylisting}
\begin{verbatim}
                0   1   2   3   4   5   6   7
              +---+---+---+---+---+---+---+---+
              | 1 | 1 | 1 | 0 |    EID    |NH |
              +---+---+---+---+---+---+---+---+
\end{verbatim}
\end{mylisting}
\caption{IPv6 Extension Header Encoding}\label{fig:ipv6.ext.header}
\end{figure}

@@

\nocite{ieee802.15.4}
\nocite{eui64}
\nocite{rfc780}
\nocite{rfc2460}
\nocite{rfc2464}
\nocite{rfc2474}
\nocite{rfc3168}
\nocite{rfc3775}
\nocite{rfc4191}
\nocite{rfc4291}
\nocite{rfc4443}
\nocite{rfc4861}
\nocite{rfc4862}
\nocite{rfc4919}
\nocite{rfc4944}
\nocite{draft-usecases-05}
\nocite{draft-hc-06}
\nocite{draft-nd-07}
\nocite{draft-routing-04}
\nocite{draft-protocols-07}
\nocite{draft-rpl-04}
\nocite{draft-routing-metrics-04}
